---
title: "Batch correction"
author: "NadineBestard"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 80
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set-up

```{r set-up, message=FALSE, warning=FALSE}
library(here) # for reproducible paths
library(SingleCellExperiment)
library(scater) # For qc and visualisation
library(scran) # For normalisation
library(ggplot2) # To add titles to plots
library(batchelor) # Batch correction
```

```{r project}
project <- "fire-mice"
```

# Batch correction with MNN

Combat was shown to outperform other batch correction methods for simple batch
correction (Buttner et. al, 2019). However, this will also regress other 
biological differences that are not well balanced between batches. Integration 
techniques account for this fact, with the downside that it can lead to
over-correction due to increased degrees of freedom of these non-linear methods. 

We use a merge tree, useful for merging together batches that are known to be 
more closely related (same tissue) before attempting difficult merges involving more dissimilar
batches.

```{r import}

```

```{r batch-correct}
if (!(file.exists(
  here("processed", project,  "sce_corrected.RDS")
))) {
sce <- readRDS(here("processed", project,  "sce_dimred_01.RDS"))
set.seed(100)
sce <- correctExperiments(sce,
  batch = factor(sce$sex),
  subset.row = rowSubset(sce),
  correct.all=TRUE,
  PARAM = FastMnnParam(
  d = 25,
  prop.k=0.05
  )
)
} else {
   sce <- readRDS(here("processed", project,  "sce_corrected.RDS"))
}
```


```{r tsne}
if (!(file.exists(
  here("processed", project,  "sce_corrected.RDS")
))) {
#keep the previous dimensional reduction just in case
reducedDim(sce, "TSNE_uncorrected") <- reducedDim(sce, "TSNE")
set.seed(100)
sce <- runTSNE(sce,  dimred="corrected")
}
# plot uit
plotReducedDim(sce, colour_by= "chip", point_size=0.1, dimred = "TSNE_uncorrected") + 
      ggtitle("TSNE dimensional reduction before correction")
plotReducedDim(sce, colour_by= "chip", point_size=0.1, dimred = "TSNE") + 
      ggtitle("TSNE dimensional reduction corrected")
```

One useful diagnostic is the proportion of variance within each batch that is 
lost during MNN correction. Specifically, this refers to the within-batch variance 
that is removed during orthogonalization with respect to the average correction 
vector at each merge step. This is in the metadata, which contains a matrix of 
the variance lost in each batch (column) at each merge step (row).

Large proportions of lost variance (>10%) suggest that correction is removing
genuine biological heterogeneity. This would occur due to violations of the 
assumption of orthogonality between the batch effect and the biological subspace.
In this case, the proportion of lost variance is small, indicating that 
non-orthogonality is not a major concern.

```{r}
metadata(sce)$merge.info$lost.var
```

```{r}
if (!(file.exists(
  here("processed", project,  "sce_corrected.RDS")
))) {  saveRDS(sce, here("processed", project,  "sce_corrected.RDS"))}
```


## Session Info

<details>

<summary>

Click to expand

</summary>

```{r session-info}
sessionInfo()
```

</details>
