---
title: "Cluster quality control"
author: "Nadine Bestard"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### set-up

```{r set-up, message=FALSE, warning=FALSE}
library(SingleCellExperiment)
library(here) # reproducible paths
library(scater) # plot reduced dims
library(tibble)  # for `rownames_to_column`
library(scDblFinder) # detect doublets
library(dplyr) # df filtering
```

```{r project}
project <- "fire-mice"
```

```{r load}
if (!file.exists(here("processed", project, "sce_anno_01.RDS"))) {
  sce <- readRDS(here("processed", project, "sce_clusters_01.RDS"))
}else{
  sce <- readRDS(here("processed", project, "sce_anno_01.RDS"))
}
```

```{r colours}
source(here("src/colours.R"))
```

### Introduction

As in [publication] we will perform a cluster QC to remove clusters of poorer
quality. This will be assessed by the number of UMI counts, the mitochondrial
percentage, doublet analysis, ribosomal genes and the number of samples that
contribute to each cluster. Moreover we will keep in mind our experimental
groups in order to ensure biological effects are not being lost.

```{r}
plotTSNE(sce, colour_by = "clusters_named", point_size=0.5,
          text_by = "clusters_named", text_size = 3) +
  scale_color_manual(values = cols) 
```

### Number of molecules per cluster

```{r}
plotTSNE(sce, colour_by = "total") + 
  ggtitle("Total number of umi counts per cell")
plotTSNE(sce, colour_by = "detected") +
  ggtitle("Detected number of genes per cell")
```

Lower values of umi counts and detected genes can be associated to lower quality
cells. Cells can also have lower expressed genes due to their biological state
or celltype.

```{r}
pct_cells<- 50
min_umi  <- 4000
```

Select clusters with `r pct_cells` % cells having less than `r min_umi`umi
counts.

```{r}
umi_df <- as.data.frame(cbind(sce$total, as.character(sce$clusters_named)))
colnames(umi_df) <- c("umi", "clusters_named")

# caculate the percentatges of cells that are "low umi"
umi_df <- 
  umi_df %>% 
  mutate(low_umi = as.numeric(umi) < min_umi) %>% 
  group_by(clusters_named) %>% 
  summarise(n_low_umi = sum(low_umi), n_cells = n()) %>%
  mutate(pct_cell_low_umi = (n_low_umi/n_cells)*100) 
umi_df %>% arrange(desc(pct_cell_low_umi))

# Filter the clusters that have a high proportion of "low umi"
low_umi_clusters <- 
  umi_df %>% 
  filter(pct_cell_low_umi > pct_cells) %>% 
  .$clusters_named
```

The clusters flagged are `r low_umi_clusters`. Mural cells is at the edge of
being flagged, however this could be due to the celltype having lower counts,
while the others are clusters of cells from celltypes already represented in
other healthier clusters. We will therefore not delete that cluster.

### Mithocondrial genes

High mithocondrial genes is associated with stressed, lower quality, cells.

```{r}
plotTSNE(sce, colour_by = "subsets_mt_percent") +#, text_by = "clusters_named") + 
  ggtitle("Percentatge mithocondrial genes")
```

```{r}
pct_cells<- 50
pct_mt  <- 5
```

Select clusters with `r pct_cells` % cells having more than `r pct_mt`%
mithocondrial genes.

```{r}
mt_df <- as.data.frame(cbind(sce$subsets_mt_percent, as.character(sce$clusters_named)))
colnames(mt_df) <- c("subsets_mt_percent", "clusters_named")

mt_df <- 
  mt_df %>% 
  mutate(high_pct = as.numeric(subsets_mt_percent) > pct_mt) %>% 
  group_by(clusters_named) %>% 
  summarise(n_high_mt = sum(high_pct), n_cells = n()) %>% 
  mutate(pct_cell_high_mt = (n_high_mt/n_cells)*100) %>%  
  arrange(desc(pct_cell_high_mt))
mt_df
high_mt_clusters <- 
  mt_df %>% 
  filter(pct_cell_high_mt > pct_cells) %>% 
  .$clusters_named

```

The clusters flagged are `r high_mt_clusters`.

### Ribosomal genes

To visualise the ribosomal content in the whole dataset we plotted the cells
according to their ribosomal content. High ribosomal content in one cluster,
that expresses a mix profile, could indicate that the cells are clustering based
on ribosomal gene content.

```{r}
# the cluster 
# save ribosomal genes
is_ribo <- grepl("^Rp[sl]", rownames(sce))
# recompute the quality metrics, subseting the ribosomal genes
ribo_qc <- perCellQCMetrics(sce, subsets = list(ribo = is_ribo))
# we are only interested in ribo
sce$subsets_ribo_percent <- ribo_qc$subsets_ribo_percent
sce$subsets_ribo_sum <- ribo_qc$subsets_ribo_sum
sce$subsets_ribo_detected <- ribo_qc$subsets_ribo_detected

```

```{r}
plotTSNE(sce, colour_by = "subsets_ribo_percent") + 
  ggtitle("Percentatge ribosomal genes")
```

### Number of cells from each sample

How many mice contribute to each cluster?

```{r cells-per-sample}
# create a summary per sample
sum_per_mouse_cluster <- table(sce$Sample, sce$clusters_named )
as.data.frame.matrix(sum_per_mouse_cluster)
```

```{r}
# count how many cells from each gnt group  there are per cluster
sum_per_smp_cluster <- table(sce$Sample, sce$clusters_named )

# normalise per cluster, looking how the KO and WT are distributed
# across the clusters, to give to both groups the same weight
prop_per_cluster <- prop.table(sum_per_smp_cluster, margin = 1)

# calculate the proportions for each cluster
prop_per_smp <- round(prop.table(prop_per_cluster , margin = 2 )*100, 2)

# Display
prop_per_smp<- as.data.frame(prop_per_smp)
colnames(prop_per_smp) <- c("Sample", "cluster", "Proportion")

ggplot(data = prop_per_smp, aes(x = cluster, y = Proportion, fill = Sample)) +
  geom_bar(position = "fill", stat = "identity") + theme(axis.text.x = element_text(angle = 45, hjust=1, vjust = 1)) +
  scale_fill_manual(values = brewer.paired(12)) 
```

### Clusters of doublet cells

Detection of clusters formed by doublets/multiplets (i.e. multiple cells
captured within the same droplet or reaction volume). The function test each
cluster against the null hypothesis that it *does* consist of doublets. The tool
consider every possible triplet of clusters consisting of a query cluster and
two putative "source" clusters. Under the null hypothesis that the query
consists of doublets from the two sources, it computes the number of genes
(`num.de`) that are differentially expressed in the same direction in the query
cluster compared to both of the source clusters. Such genes would be unique
markers for the query cluster and provide evidence against the null hypothesis.

-   Clusters with the few unique genes (`num.de`) are more likely to be composed
    of doublets.

-   Potential doublets should also have larger library size than the cells from
    the source clusters, resulting in `lib.size` ratios bellow unity.

```{r}
if(!file.exists(here("outs",project,  "doublet_clusters_res0.3.csv"))){
  res_dbl <- findDoubletClusters(sce, sce$clusters_named)
  res_dbl <- res_dbl %>% 
    as.data.frame() %>% 
    rownames_to_column("cluster")
  write.csv(res_dbl, here("outs", project,  "doublet_clusters_res0.3.csv"))
}else{
  res_dbl <- read.csv(here("outs", project,  "doublet_clusters_res0.3.csv"), row.names = 1)
}
res_dbl
```

All clusters have a significant p-value to reject the null hypothesis that they
do consist of doublets. The clusters with fewer DE genes with the potential
sources are microglia clusters, likely formed by microglia engulfing other cell
types. The next clusters are compositions of poor-quality clusters, making it
hard to untangle if these are doublets. A first clusterQC was made keeping the
potential doublets (commit 15a39b03), just deleting the clusters flagged for
other reasons, and then doubletFinder was run again to elucidate the doublets.
Results below:

Astro-OPCs will not be kept. Astro-Oligo may contain some doublets too but the
cluster will be kept due to evidence of astro-oligo populations in tissue.

+-------+-------+-------+-------+-------+-------+-------+-------+-------+-------+
| cl    | so    | so    | n     | medi  | best  | p.    | lib.  | lib.  | prop  |
| uster | urce1 | urce2 | um.de | an.de |       | value | size1 | size2 |       |
+-------+-------+-------+-------+-------+-------+-------+-------+-------+-------+
| t     | tOPC1 | tA    | 344   | 5214  | Apoe  | 2.4   | 0.76  | 0.40  | 0.01  |
| Astro |       | stro1 |       |       |       | 2E-43 | 93783 | 58445 | 33811 |
| _OPCs |       |       |       |       |       |       |       |       |       |
+-------+-------+-------+-------+-------+-------+-------+-------+-------+-------+
| tmO  | tm   | tA   | 839  | 3975 | Rna  | 1.1  | 0.79 | 0.43 | 0.01 |
| ligo_ | Oligo | stro1 |       |       | set2a | 7E-21 | 73161 | 48667 | 06727 |
| Astro |       |       |       |       |       |       |       |       |       |
+-------+-------+-------+-------+-------+-------+-------+-------+-------+-------+
| t     | tRibo | tmN   | 1627  | 4490  | Vtn   | 0.0   | 1.86  | 3.06  | 0.01  |
| Mural | somal | euron |       |       |       | 0E+00 | 14346 | 18305 | 48959 |
| cells | _High |       |       |       |       |       |       |       |       |
|       | _Micr |       |       |       |       |       |       |       |       |
|       | oglia |       |       |       |       |       |       |       |       |
+-------+-------+-------+-------+-------+-------+-------+-------+-------+-------+
| tE    | tRibo | tmN   | 1836  | 4936  | Ly6c1 | 0.0   | 1.69  | 2.7   | 0.01  |
| ndoth | somal | euron |       |       |       | 0E+00 | 23566 | 83718 | 83387 |
| elial | _High |       |       |       |       |       |       |       |       |
|       | _Micr |       |       |       |       |       |       |       |       |
|       | oglia |       |       |       |       |       |       |       |       |
+-------+-------+-------+-------+-------+-------+-------+-------+-------+-------+

```{r}
doublet <- c("tOPC_Microglia", "tAstro_Microglia", "tAstro_OPCs")
```

<!-- ### Control vs fire mice -->

We want to have a closer look at the clusters that do have a difference between
the knockout and the wild type before deleting the clusters.

```{r fig.height=15, fig.width=10, message=FALSE}
# divide the two objects
sce_ko <- sce[,sce$genotype == "KO"]
sce_ctr <- sce[,sce$genotype == "WT"]
# plot them side by side
gridExtra::grid.arrange(
plotTSNE(sce_ko, colour_by = "clusters_named", point_size=0.5,
         point_alpha = 0.3, text_by = "clusters_named", text_size = 3) +
  scale_color_manual(values = cols) +
  ggtitle("fire mice"),
plotTSNE(sce_ctr, colour_by = "clusters_named", point_size=0.5,
         point_alpha = 0.3, text_by = "clusters_named", text_size = 3) +
  scale_color_manual(values = cols) +
  ggtitle("control"),
ncol = 1
)
```

#### Proportion KO-WT

The poor quality cluster that are more abundant in KO than WT are coming mostly
from one single sample, they will therefore not be "protected" and will be
discarded.

<!-- In order to visualise the proportions from KO and WT for each cluster, we do not -->

<!-- take in consideration the microglia clusters, as these are only present in the -->

<!-- control, and we normalise per number of cells per cluster. -->

<!-- ```{r} -->

<!-- # delete the microglia -->

<!-- sce_no_mc <- sce[,!grepl("Microglia", sce$clusters_named)] -->

<!-- # count how many cells from each gnt group  there are per cluster -->

<!-- sum_per_gnt_cluster_no_mc <- table(sce_no_mc$genotype, sce_no_mc$clusters_named ) -->

<!-- # normalise per cluster, looking how the KO and WT are distributed -->

<!-- # across the clusters, to give to both groups the same weight -->

<!-- prop_per_cluster_no_mc <- prop.table(sum_per_gnt_cluster_no_mc, margin = 1) -->

<!-- # calculate the proportions for each cluster -->

<!-- prop_per_gnt_no_mc <- round(prop.table(prop_per_cluster_no_mc , margin = 2 )*100, 2) -->

<!-- # Display -->

<!-- prop_per_gnt_no_mc<- as.data.frame(prop_per_gnt_no_mc) -->

<!-- colnames(prop_per_gnt_no_mc) <- c("Genotype", "cluster", "Proportion") -->

<!-- prop_per_gnt_no_mc %>% -->

<!--   filter(Genotype == "KO") %>% -->

<!--   arrange(desc(Proportion)) -->

<!-- ``` -->

<!-- visualise in a plot -->

<!-- ```{r plot} -->

<!-- ggplot(data = prop_per_gnt_no_mc, aes(x = cluster, y = Proportion, fill = Genotype)) + -->

<!--   geom_bar(position = "fill", stat = "identity") + theme(axis.text.x = element_text(angle = 45, hjust=1, vjust = 1)) + -->

<!--   scale_fill_manual(values = col_wt_ko) -->

<!-- ``` -->

<!-- ```{r} -->

<!-- ko_pct <- 65 -->

<!-- ``` -->

<!-- ```{r} -->

<!-- difference_KO_WT <- -->

<!--   prop_per_gnt_no_mc %>% -->

<!--   arrange(desc(Proportion)) %>% -->

<!-- # the microglia clusters will be the top differences, but this is not -->

<!--     # something we are specially interested -->

<!--   filter(!(cluster %in% c("Microglia1", "Microglia2"))) %>% -->

<!--     filter(Genotype == "KO") %>% -->

<!--     # select the "interesting" clusters -->

<!--   filter(Proportion > ko_pct) %>% -->

<!--   .$cluster -->

<!-- ``` -->

<!-- ```{r} -->

<!-- gt_ko_pct <- unique(sce$clusters_named[sce$clusters_named %in% c( high_mt_clusters, low_umi_clusters, doublet) & -->

<!--     (sce$clusters_named %in% difference_KO_WT)]) -->

<!-- #gt_ko_pct -->

<!-- ``` -->

<!-- There are r length(gt_ko_pct)` clusters flagged as poor quality and also more -->

<!-- abundant (greater than r ko_pct` %) in the KO animals, these are r gt_ko_pct`. -->

### Cluster QC

the clusters delted from downstream analysis are highlighted bellow, and are
composed by clusters with:

-   low umi: more than 50% of cells having less than 4000 umi counts. These are:

`r low_umi_clusters`

-   high mt : majority of cells having more than 5 % mitochondrial genes

`r high_mt_clusters`

-   doublet: clusters composed by cells that express markers from different cell
    types.

`r doublet`

<!-- -   artefacts coming from few samples: -->

<!-- r low_n_mice` -->

<!-- -   Clusters with different abundance between KO and WT were kept, as they could be different due to the biological condition. -->

<!-- r gt_ko_pct` -->

```{r filterout}
sce$filter_out <- sce$clusters_named %in% c( high_mt_clusters, low_umi_clusters, doublet) 
filter_out <- sum(sce$filter_out)
plotTSNE(sce, colour_by = "filter_out", point_size=0.5,
         point_alpha = 0.3, text_by = "clusters_named", text_size = 3) 
#plotTSNE(sce, colour_by = "filter_out", point_size=0.5,
#         point_alpha = 0.3, text_by = "celltype", text_size = 3)
```

```{r save, eval=TRUE, include=TRUE}
if (!file.exists(here("processed", project, "sce_clusterQC.RDS"))){
 sce <- sce[, sce$filter_out == FALSE]
 saveRDS(sce, here("processed", project, "sce_clusterQC.RDS"))
}


```

This filters `r filter_out` cells, leaving an object with
`r sum(sce$filter_out == FALSE)`cells.

## Session Info

<details>

<summary>Click to expand</summary>

```{r session-info}
sessionInfo()
```

</details>
